# Example solutions for exercises {#sec-example-solutions}

```{r}
#| label: setup
#| echo: false
#| results: asis

library(rebook)
chapterPreamble()

knitr::opts_chunk$set(warning = FALSE, message = FALSE, output = FALSE)
```

## Introduction

These example solutions are related to exercises in
[@sec-microbiome-in-bioconductor] and [@sec-ecosystem].

```{r}
#| label: intro
#| code-fold: true
#| code-summary: "Show the solution"

```

## Data containers

These example solutions are related to exercises in [@sec-containers].

```{r}
#| label: containers
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Import

These example solutions are related to exercises in [@sec-import].

```{r}
#| label: import
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Convert & export

These example solutions are related to exercises in [@sec-convert].

```{r}
#| label: convert
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Taxonomic information

These example solutions are related to exercises in
[@sec-taxonomic-information].

```{r}
#| label: taxonomy
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Data wrangling

These example solutions are related to exercises in [@sec-datawrangling].

```{r}
#| label: wrangling
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Agglomeration

These example solutions are related to exercises in [@sec-agglomeration].

```{r}
#| label: agglomeration
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Transformation

These example solutions are related to exercises in [@sec-assay-transform].

```{r}
#| label: transformation
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Exploration & quality control

These example solutions are related to exercises in [@sec-quality-control].

```{r}
#| label: qc
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Community composition

These example solutions are related to exercises in [@sec-microbiome-community].

```{r}
#| label: composition
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Alpha Diversity

These example solutions are related to exercises in [@sec-alpha-diversity].

```{r}
#| label: alpha
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Community similarity

These exercises are related to exercises in [@sec-community-similarity].

```{r}
#| label: beta
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Community typing

These example solutions are related to exercises in [@sec-clustering].

```{r}
#| label: cluster
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Differential abundance

These example solutions are related to exercises in
[@sec-differential-abundance].

```{r}
#| label: daa
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Correlation

These example solutions are related to exercises in [@sec-correlation].

```{r}
#| label: correlation
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Mediation

These example solutions are related to exercises in [@sec-mediation].

```{r}
#| label: mediation
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Network learning & analysis

These example solutions are related to exercises in [@sec-network-learning].

```{r}
#| label: network1
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Network comparison

These example solutions are related to exercises in [@sec-network-comparison].

```{r}
#| label: network2
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Cross-association

These example solutions are related to exercises in [@sec-cross-correlation].

```{r}
#| label: cross-corr
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Ordination-based multiassay analysis

These example solutions are related to exercises in
[@sec-multiassay_ordination].

```{r}
#| label: mofa
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Multi-omics prediction and classification

These example solutions are related to exercises in
[@sec-multi-omics-integration].

```{r}
#| label: integratedlearner
#| code-fold: true
#| code-summary: "Show the solution"
    
```

## Machine learning

These example solutions are related to exercises in [@sec-machine_learning].

```{r}
#| label: ml
#| code-fold: true
#| code-summary: "Show the solution"
    
```
