# Exploration & quality control {#sec-quality-control}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

This chapter focuses on the quality control (QC) and exploration of
microbiome data and establishes commonly used descriptive
summaries. Familiarizing yourself with the peculiarities of a given dataset is
essential for data analysis and model building to make justified conclusions.

The dataset should not suffer from severe technical biases, and you
should at least be aware of potential challenges, such as outliers,
biases, unexpected patterns, and so forth. Standard summaries and
visualizations can help, and the rest comes with experience. Moreover,
exploration and QC often entail iterative processes.

There are available guidelines for QC, for instance, [@Zuur2010].
However, it should be noted that one should not follow any protocols strictly
but rather use them as a template and customize them to one's dataset. The goal
of QC is not only to improve quality of data, but also to understand it and its limitations. Poor data quality leads to poor results which is illustrated by a
common expression in computer science:

> _"Garbage in, garbage out"_

Below we download a dataset from `microbiomeDataSets`.
The dataset contains 16S data from the gut microbiome of baboons.

```{r}
#| label: load_data

library(microbiomeDataSets)
tse <- baboongut()
```

## Summarize data {#sec-qc-summarize}

When you first get your hands dirty with the data, the first step is to
summarize it and get a sense of what kind of data you're dealing with. Printing
`TreeSE` already provides useful information about the dataset's dimensions,
such as the number of samples and taxa.

```{r}
#| label: explore1

tse
```

The dataset includes `r nrow(tse)` taxa identified from `r ncol(tse)` samples.
`mia` provides the `summary()` function for `TreeSE` objects which returns 
the summary of counts for all samples and features including measures of 
central tendency.

```{r}
#| label: explore2

library(mia)

# Calculate summary tables
summary(tse, assay.type = "counts")
```

The returned tables show that samples exhibit lots of variance in library sizes.
Moreover, we can observe that there are no singletons in the dataset
(library size and singletons are discussed in more detail in
[@sec-qc-outliers]).

Another type of summary can be generated using the `summarizeDominance()`
function. This function returns a table displaying both the absolute and
relative abundance of each taxon &mdash; that is, how many times a taxon was
detected in the dataset and the proportion of samples in which it was
identified. Below, we create a summary table for genera.

```{r}
#| label: explore3

df <- summarizeDominance(tse, rank = "Genus")
df
```

Based on the summary table, `r df[["dominant_taxa"]][[1L]]` seems to be
highly presented in the baboon gut.

`mia` also provides other functions to summarize the dataset. For example, you
can retrieve unique, top, prevalent, or rare taxa using `getUnique()`,
`getTop()`, `getPrevalent()`, and `getRare()`, respectively.

Let’s first check which phyla are included in the dataset.

```{r}
#| label: explore4

uniq <- getUnique(tse, rank = "Phylum")
uniq
```

There are `r length(uniq)` phyla present in the dataset.

Next, we might be interested on assessing which features are the most abundance.
This can be done by utilizing `getTop()` function. Below, we pick top 10 taxa
which is determined based on median abundance.

```{r}
#| label: explore5

# Pick the top taxa
top_features <- getTop(tse, method = "median", top = 10)
top_features
```

These ten features, have the highest median abundance across all samples.
`getPrevalent()` differs from `getTop()` so that it retrieves which taxa
exceeds the specified prevalence and detection thresholds. Here we determine
prevalent genera.

```{r}
#| label: explore6

prev <- getPrevalent(tse, rank = "Genus", prevalence = 0.2, detection = 0)
prev |> head()
```

Out of `r library(mia); nrow(agglomerateByRank(tse, "Genus"))` genera,
`r length(prev)` of them are detected to be prevalent, meaning they are found
in a sufficient number of samples with sufficiently high abundance.

`getRare()` function is counterpart of `getPrevalent()`. It returns those taxa
that do not exceed the thresholds. 

```{r}
#| label: explore7

rare <- getRare(tse, rank = "Genus", prevalence = 0.2, detection = 0)
rare |> head()
```

It returns `r length(rare)` taxa which is expected as it should match with
number of genera after subtracting the prevalent ones.

## Outliers, singletons and contaminants {#sec-qc-outliers}

Outliers are observations that deviate significantly from the rest of the data.
Researchers should be cautious when handling outliers, as what appears to be an
outlier may actually result from a valid biological mechanism. This is
especially relevant in the microbiome field, where sequenced counts often
exhibit seemingly irregular variation. In such cases, transformations (see
[@sec-assay-transform]) are typically the preferred approach for handling these
observations. However, if an observation or sample is significantly affected by
measurement error, removing it may be a reasonable option. The approach to
dealing with outliers in each dataset requires careful consideration by the
researcher.

Singletons are sequences that appears only once in a dataset, meaning it is
observed in just a single sequencing read. Often rare features are removed from
the data as they usually represent sequencing artifacts. See [@sec-subset_prev]
and [@sec-agglomerate_prev] for more details on prevalence filtering and
agglomeration.

### Library size

Library size refers to the total number of counts found in a single sample.
The returned tables in [@sec-qc-summarize] showed that samples exhibit lots of
variance in library sizes. We can then visualize the library sizes. The total
counts per sample can be calculated using `addPerCellQC()` from the `scater`
package.

```{r}
#| label: outlier2

library(scater)
library(knitr)

# Calculate and add total counts to `colData`
tse <- addPerCellQC(tse)
```

The results are stored in `colData`. We can then visualize these results with a
violin plot and histogram.

```{r}
#| label: outlier3

library(patchwork)

p1 <- plotColData(tse, x = "sex", y = "total", colour_by = "age")
p2 <- plotHistogram(tse, col.var = "total")

p1 + p2
```

The distribution of library size is right-skewed. Most of the samples follow
normal distribution while some of the samples deviates from this. This might be
caused by technical variations or biological factors. 

Nevertheless, the most important check is to ensure that the sampling depth is
sufficient. In case of insufficient sampling depth, one might consider filtering
the data based on library size ([@sec-subset-library-size]).

To control uneven sampling depths, one should apply data transformation or apply
rarefaction. These both approaches are discussed in [@sec-assay-transform].

### Contaminant sequences

Samples might be contaminated with exogenous sequences. The impact of
each contaminant can be estimated based on its frequencies and
concentrations across the samples.

The following
[decontam functions](https://microbiome.github.io/mia/reference/isContaminant.html)
are based on [@davis2018simple] and support such functionality:

* `isContaminant()`, `isNotContaminant()`
* `addContaminantQC()`, `addNotContaminantQC()`

Contaminations can be detected using two main approaches: frequency-based and
prevalence-based testing. In frequency-based testing, the user must provide the
DNA concentration of samples. The abundance of features is then compared to the
DNA concentration, as contaminants are expected to show an inverse
relationship &mdash; they make up a larger fraction in low-DNA samples and 
smaller fraction in high-DNA samples.

In the prevalence-based approach, sequence prevalence is compared between true
biological samples and control samples. This method assumes that contaminants
are more prevalent in negative controls, as these lack true biological DNA and
primarily contain background noise from contamination sources.

```{r}
#| label: replace_missing_values
#| echo: false

# These is a bug in decontam. If there are missing values, it leads to an error
# in input check. This is why we replace missing values with imputed values.
# This chunk can be removed when this PR is merged:
# https://github.com/benjjneb/decontam/pull/162
avg_conc <- mean(tse[["post_pcr_dna_ng"]], na.rm = TRUE)
colData(tse)[ is.na(tse[["post_pcr_dna_ng"]]), "post_pcr_dna_ng"] <- avg_conc
```

The dataset contains DNA concentration recorded in the "post_pcr_dna_ng" column
in the sample metadata.
This information can be used for frequency-based contamination identification.

```{r}
#| label: show_metadata

library(knitr)

colData(tse) |> head() |> kable()
```

Now we can detect contaminant sequences. We run `addContaminantQC()` which
adds results to `rowData`. 

```{r}
#| label: detect_contaminant

library(mia)

tse <- addContaminantQC(tse, concentration = "post_pcr_dna_ng")

rowData(tse) |> head() |> kable()
```


The method performs statistical tests to identify contaminants. By default, it
uses a 0.1 probability threshold for identifying contaminants, which can be
adjusted as needed. In this case,
`r (sum(rowData(tse)[["contaminant_contaminant"]])/nrow(tse))*100`% of
sequences were detected to be contaminants. We can then filter out these
features from the data.

```{r}
#| label: filter_contaminants

tse <- tse[ !rowData(tse)[["contaminant"]], ]
```

As an example, we also demonstrate how to apply the prevalence-based approach.
Note that the data used here is arbitrary, and in practice, you should use real
control sample information.

First, we add arbitrary control samples:

```{r}
#| label: add_control_info

control <- rep(FALSE, ncol(tse))
control[ sample(seq_len(ncol(tse)), 1) ] <- TRUE
tse[["control"]] <- control
```

Next, we perform the analysis. Note that we could have applied both
frequency-based and prevalence-based methods simultaneously by specifying the
`method` parameter in the previous step.

```{r}
#| label: contaminant_prevalence

not_contaminant <- isNotContaminant(tse, control = "control", detailed = FALSE)
not_contaminant |> head()
```

To identify non-contaminant sequences with prevalence approach, a threshold of
0.5 is used, by default. As noted in the previous step, the `add*()` functions
add results to `rowData`. Here, we used function that return only the results.
By specifying `detailed = FALSE`, we obtain the results as a vector, which can
then be used for subsetting the data.  

## Data distribution

Microbiome counts data is rarely normally distributed
(see [@sec-stat-challenges]). However, many common statistical tests assume
normality and using them while violating the assumptions might lead to
incorrect conclusions. While several tests for normality exist &mdash; such as
Shapiro-Wilk test &mdash; they do not replace visual observation.

`plotHistogram()` function provides easy way to visualize the counts
distribution with a histogram.

```{r}
#| label: distribution

plotHistogram(tse, assay.type = "counts")
```

The microbiome data is typically zero-inflated, meaning that there are lots of
zeroes. Same method can also be used to visualize continuous variables from
`colData`. For categorical values, one can utilize `plotBarplot()`.

```{r}
#| label: distribution2

p1 <- plotHistogram(tse, col.var = "age") +
    labs(x = "Age")
p2 <- plotBarplot(tse, col.var = "sex") +
    labs(x = "Sex")

p1 + p2
```

### Abundance

Abundance visualization is an important data exploration
approach. `plotAbundanceDensity()` function generates a plot to visualize
the most abundant taxa along with several options.

Next, a few demonstrations are shown using the [@Lahti2014]
dataset. A jitter plot based on relative abundance data, similar to
the one presented at [@Salosensaari2021] (Supplementary Fig.1), can
be visualized as follows:

```{r}
#| label: plot_abundance

library(miaViz)

# Add relative abundances
tse <- transformAssay(tse, method = "relabundance")

plotAbundanceDensity(
    tse, layout = "jitter",
    assay.type = "relabundance",
    n = 40, point.size=1, point.shape=19,
    point.alpha = 0.1) +
    scale_x_log10(label=scales::percent)
```

The relative abundance values for the top-5 taxonomic features can be
visualized as a density plot over a log-scaled axis, with
"sex" indicated by colors:

```{r}
#| label: plot_abundance2

plotAbundanceDensity(
    tse, layout = "density",
    assay.type = "relabundance",
    n = 5, colour.by = "sex",
    point.alpha = 0.1) +
    scale_x_log10()
```

Alternatively, `scater::plotExpression()` can be used to visualize taxa with
a violin plot. Below, we visualize top-10 features, selected based on their
mean abundance.

```{r}
#| label: plot_abundance3

# Select top features
top <- getTop(tse, top = 10L, method = "mean")

plotExpression(
    tse,
    features = top,
    x = "sex",
    assay.type = "relabundance",
    point_alpha = 0.01
    ) +
    scale_y_log10()
```

### Prevalence

Prevalence quantifies the frequency of samples where certain microbes
were detected (above a given detection threshold). Prevalence can
be given as sample size (N) or percentage (unit interval).

Investigating prevalence allows you either to focus on changes which
pertain to the majority of the samples, or identify rare microbes,
which may be _conditionally abundant_ in a small number of samples.

We can plot histogram of prevalence of taxa. This would tell us whether there
are many taxa present in most of the samples or are there mostly rare
taxa. The population prevalence (frequency) at a 0.01% relative abundance
threshold (`detection = 0.1/100` and `as.relative = TRUE`) can look
like this.

```{r}
#| label: plot_prevalence

# Add prevalence of features
tse <- addPrevalence(tse, detection = 0.1/100, as.relative = TRUE)
# Plot them with a histogram
plotHistogram(tse, row.var = "prevalence")
```

Most of the taxa are present only in minority of samples with specified
abundance threshold. Similar conclusion can be made with visualization
generated with `plotPrevalentAbundance()` or `plotRowPrevalence()` functions.

```{r}
#| label: plot_prevalence2

p1 <- plotPrevalentAbundance(tse, as.relative = TRUE)
# Remove y axis text as there are so many features that one cannot read them
p2 <- plotRowPrevalence(tse, as.relative = TRUE) +
    theme(
       axis.text.y=element_blank(), 
       axis.ticks.y=element_blank()
    )

p1 + p2
```

The plots above shows that most of the taxa are present with low abundance.
However, there are couple taxa that are both high abundant and prevalent in the
dataset.

The figures also show the overall trend that most microbes are low in abundance
and occur in a limited number of samples. This is a typical pattern in
microbiome datasets, and can further visualized as follows.

```{r}
#| label: plot_prevalence3

plotPrevalence(tse, as.relative = TRUE)
```

The plot shows the relationship between microbial relative abundance and
prevalence across samples. Taxa with higher prevalence (yellow) tend to have
lower relative abundance, while less prevalent taxa (purple) can still reach
higher abundances in some cases.

To analyze how present the core taxa are in samples, we can calculate how large
proportions the core taxa present in each sample. We can then visualize this
distribution again with a histogram.

```{r}
#| label: plot_prevalence4

# Calculate the proportion of core taxa
tse <- addPrevalentAbundance(tse, prevalence = 50/100, detection = 0.1/100)
# Visualize
plotHistogram(tse, col.var = "prevalent_abundance")
```

In most of the samples, the core taxa represent over 3/4 of species. However,
there are some samples where this proportion is much lower.

## Colinearity and independence

Microbial species are rarely independent; rather, they influence each other's
abundance through complex networks of competition and symbiosis. Colinearity
occurs when information from one variable is already included in some other
variable. Modeling variables that exhibit collinearity can lead to issues such
as reduced interpretability, overfitting and incorrect estimations.

Collinearity of variables can be assessed, for instance, with correlation
heatmaps (see [@sec-correlation]) or with scatter plots. 

::: {.callout-tip icon=false}
## Exercises

**Goal:** The goal is to learn relevant function in quality control and initial
exploration of the data.

**Exercise 1: QC and exploration**

1. Load any of the example datasets mentioned in [@sec-example-data].

2. Summarize the counts with with histogram.

3. Add prevalence of taxa to `rowData`.

4. Visualize the prevalence distribution with histogram. Does the data include
many prevalent taxa or are they found only in small part of samples?

5. Add library sizes to `colData`.

6. Visualize the library size distribution with histogram. Does the sampling
depth differ a lot?

7. Visualize categorical values in `colData` with a bar plot.

8. Get the available taxonomy ranks in the data.

8. Calculate a table that summarizes the dominance of genera or any other rank.
Which taxa is present in the highest number of samples? In how many samples it
is present? What percentage of samples the number corresponds to?

9. Get the most prevalent features in specific taxonomy rank. Use counts table,
and set prevalence and detection threshold to 20% and 1, respectively.

10. Get the most abundant features based on median abundance. How this differs
from prevalent features?

11. Visualize the most prevalent features.

Useful functions:

`data()`, `plotHistogram()`, `addPrevalence()`, `rowData()`,
`addPerCellQCMetrics()`, `colData()`, `plotBarplot()`, `taxonomyRanks()`,
`summarizeDominance()`, `getPrevalent()`, `getTop`, `plotAbundanceDensity()`
:::
