# Subsetting {#sec-treese_subsetting}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

In this chapter, we explore the concept of subsetting. Subsetting or
filtering is the process of selecting specific rows or columns from a
dataset based on certain criteria. When you subset your data, you retain the
original values of the selected data points. For example, if you have a
dataset of microbial taxa and you choose to include only certain species, you
retain the individual counts for those species, while losing the information 
on other species that are not included.

Subsetting data helps to draw the focus of an analysis on particular
sets of samples and / or features. This can be useful in data cleaning 
(e.g., getting rid of non-target features or duplicate samples) or in 
dealing with large datasets, where analysing smaller subsets of data separately 
may improve performance and reduce the computational load.

Let us store `GlobalPatterns` into `tse` and check its original number of
features (rows) and samples (columns).

```{r}
#| label: load_data

# Load libraries and data
library(mia)
library(dplyr)
library(knitr)

data("GlobalPatterns", package = "mia")
tse <- GlobalPatterns

dim(tse)
```

::: {.callout-note}
## Note

When subsetting by sample, expect the number of columns to decrease. When 
subsetting by feature, expect the number of rows to decrease.
:::

## Subset by sample (column-wise)

In this particular example data several criteria can be used to subset by sample:

- origin
- sampling time
- sequencing method
- DNA / RNA barcode
- cohort

For the sake of demonstration, here we will extract a subset containing only the
samples of human origin (feces, skin or tongue). This information is stored in `SampleType` 
column within `colData(tse)` as well as in `tse`.

First, we would like to see all the possible values that `SampleType` can have
and how frequent those are:

```{r}
#| label: show_data

# Show the frequency of each value
tse$SampleType |> table() |> kable()
```

Next, we make a _logical index_ across the columns of `tse` (make sure to
leave the first index empty to include all rows) and filter for the
samples of human origin. For this, we use the information on the
samples from the sample meta data `colData(tse)`.

```{r}
#| label: filter_samples

# Subset by sample
tse_sub <- tse[ , tse$SampleType %in% c("Feces", "Skin", "Tongue")]

# Show dimensions
dim(tse_sub)
```

::: {.callout-note}
## Note

After subsetting, expect the number of columns to equal the
sum of the frequencies of the samples that you are interested
in. For instance, `ncols = Feces + Skin + Tongue = 4 + 3 + 2 = 9`.
:::

## Subset by feature (row-wise)

Similarly, here we will extract a subset containing only the taxa (features)
that belong to the phyla Actinobacteria and Chlamydiae. This information is stored in `Phylum` 
column within `rowData(tse)` as well as in `tse`. Note that subsetting by feature comes with
a few more obstacles than subsetting by sample, such as the presence of `NA` elements and the
possible need for agglomeration.

As previously, we would first like to see all the possible values that `Phylum`
can have and how frequent those are:

```{r}
#| label: show_phyla

# Show the frequency of each value
rowData(tse)$Phylum |> table() |> head() |> kable()
```

::: {.callout-note}
## Note

After subsetting, expect the number of columns to equal the
sum of the frequencies of the feature(s) that you are interested
in. For instance, `nrows = Actinobacteria + Chlamydiae = 1631 + 21 = 1652`.
:::

Depending on your research question, you might or might not need to
agglomerate the data in the first place: For example, if you want to find the
separate abundances of each and every taxon that belongs to Actinobacteria and
Chlamydiae, agglomeration is not needed. However, if you want to find the total
abundance of all taxa that belong to Actinobacteria or
Chlamydiae, agglomeration is recommended (see
[@sec-agglomeration] for details).

Next, we make a _logical index_ across the rows of `tse` (make sure to leave
the second index empty to include all columns) and filter for the
taxa that fall in either Actinobacteria or Chlamydiae group. For this,
we use the information on the features (=taxa) from the feature metadata
`rowData(tse)`.

The first term with the `%in%` operator includes all the features
of interest, whereas the second term after the AND operator `&`
filters out all features that have an `NA` in place of the phylum variable.

```{r}
#| label: subset_features

# Subset by feature
selected <- rowData(tse)$Phylum %in% c("Actinobacteria", "Chlamydiae") &
  !is.na(rowData(tse)$Phylum)
tse_sub <- tse[selected, ]

# Show dimensions
dim(tse_sub)
```

## Subset by samples and features

Finally, we can subset data by sample and feature at once. The
resulting subset contains all the samples of human origin and all the
features of phyla Actinobacteria or Chlamydiae.

```{r}
#| label: subset_samples_and_feat

# Subset by sample and feature and remove NAs
selected_rows <- rowData(tse)$Phylum %in% c("Actinobacteria", "Chlamydiae") &
      !is.na(rowData(tse)$Phylum)
selected_cols <- tse$SampleType %in% c("Feces", "Skin", "Tongue")
tse_sub <- tse[selected_rows, selected_cols]

# Show dimensions
dim(tse_sub)
```

::: {.callout-note}
## Note

The dimensions of `tse_sub` are consistent with the previous subsets
(9 columns filtered by sample and 1652
rows filtered by feature).
:::

If a study was to consider and quantify the presence of Actinobacteria
as well as Chlamydiae in different sites of the human body,
`tse_sub` might be a suitable subset to start with.

## Filtering based on library size {#sec-subset-library-size}

As a preprocessing step, one might want to remove samples below
certain library size, i.e., total number of counts.

```{r}
#| label: subset_libsize

# Calculate the total number of counts in each sample in this data.
library(scuttle)
tse <- addPerCellQCMetrics(tse)
tse$total |> head()
# Inspect distribution of library sizes among samples:
summary(tse$total)

# Remove samples that have a library size smaller than 100 000
tse_sub <- tse[ , tse$total > 100000]
tse_sub
```

Additionally, in some cases, such as after subsetting to filter some features away,
data might contain "empty" samples that do not contain any of the
features present in the remaining dataset. To focus only samples containing 
sufficient information, we might want to remove these instances. 

In this example, we are first interested only those features (=taxa) that belong to 
Species _Achromatiumoxaliferum_.

```{r}
#| label: select_archaea

ind <- rowData(tse)$Species == "Achromatiumoxaliferum"
ind[is.na(ind)] <- FALSE
tse_sub <- tse[ind, ]
```

Then we can again calculate the total number of counts in each sample in this subsetted data.

```{r}
#| label: subset_empty

library(scuttle)
tse_sub <- addPerCellQCMetrics(tse_sub)
# List total counts of each sample
tse_sub$total |> head()
```

We can see now that, after our feature-based subsetting step, many samples do not include any counts. 
These are samples in the original data, that had no counts of _Achromatiumoxaliferum_.
We can remove these empty samples from our data subset.

```{r}
#| label: subset_empty2

# Remove samples that do not contain any bacteria
tse_sub <- tse_sub[ , tse_sub$total != 0]
tse_sub
```

We have now subsetted the dataset so that it only includes selected features and only samples 
containing these selected features. A similar approach can also be applied to filter features
based on the samples they appear in, ensuring both dimensions of the data are
relevant to your analysis.

## Filtering out zero-variance features

In some datasets, certain features remain constant across all samples — that is, they show
no variance. If our goal is to study or model microbial differences between
groups, these zero-variance features hold no value. For a feature to reflect
differences between groups, it must exhibit variability.

By removing these invariant features, we can sharpen our focus on the more
informative features. This not only reduces the number of comparisons but can also
help our models to learn from meaningful data without additional noise.

To filter out invariant features, we begin by calculating the standard deviation across all features.
Then we visualize the distribution of these variances as a histogram.

```{r}
#| label: standard_deviation

# Calculate standard deviation of each feature's counts across all samples and add this as an additional
column in the feature data (`rowData`)
rowData(tse)[["sd"]] <- rowSds(assay(tse, "counts"))
# Plot distribution of standard deviations across features 
# Note: we are plotting the standard deviations in a log scale to see the breadth of their variance better
hist(log(rowData(tse)[["sd"]]))
```

Based on the histogram of feature count variances, we can decide a sensible threshold
for filtering. Here, using a threshold of 0 (i.e., log-transformed standard deviation = 0)
would effectively remove a large set of features that show no variance (i.e., standard deviation = 1).

It's important to note that the data is on a log scale, meaning that a
log-transformed value of 1 corresponds to 0 (i.e., log(1) = 0). This ensures
that features with zero variance are correctly identified and filtered out.

```{r}
#| label: filter_sd

selected <- rowData(tse)[["sd"]] > 1
tse_sub <- tse[selected, ]
tse_sub
```

After filtering, the dataset includes the `r nrow(tse_sub)` features comapred to previous
`r nrow(tse)`. These are the features that vary at all. Note that
this threshold choice depends on your research question, and this was a rather
conservative choice retaining most of the features. We could have also filtered features based
on having variance above a certain non-zero threshold.

## Subset based on prevalence {#sec-subset_prev}

Sometimes we might want to subset the data based on feature prevalence, that is
the commonness/rarity of a feature across the samples. We can do this by using 
`subsetByPrevalent()`, which filters taxa that exceed a specified prevalence threshold. 
This can be useful for example if we wish to remove rare taxa in our data, as these
may be sequencing artefacts rather than real taxa.

Conversely, `subsetByRare()` allows us to retain only features _below_ the prevalence threshold, 
enabling a focus on rare features within the dataset.

Here, we apply a prevalence threshold of 10% and a detection threshold of 1.
This means a feature is considered "detected" if it has at least one count in a sample, and
prevalent if it is detected in at least 10% of the samples. The function
`subsetByPrevalent()` retains the features more prevalent than this, while `subsetByRare()`
keeps the features that are less prevalent (rarer) than this.

```{r}
#| label: subset_by_rare

tse_sub <- subsetByRare(tse, rank = "Genus", prevalence = 0.1, detection = 1)
tse_sub
```

In some cases, agglomerating based on prevalence can provide more meaningful
insights. This process is illustrated in [@sec-agglomerate_prev].
