# Transformation {#sec-assay-transform}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

Data transformations are common in (microbial) ecology [@Legendre2001]
and used to improve compatibility with assumptions related to specific
statistical methods, mitigate biases, enhance the comparability of
samples, or features, or to obtain more interpretable values.

Examples include the logarithmic transformation, the calculation of
relative abundances (percentages), and compositionality-aware
transformations such as the centered log-ratio transformation (clr).

## Characteristics of microbiome data {#sec-stat-challenges}

Microbiome sequencing data has unique characteristics that must be addressed;
otherwise, incorrect decisions might be made based on the results.
Specifically, microbiome sequencing data is
characterized by high variability, zero-inflation and compositionality.
**High variability** expresses that abundance of taxa often varies by several
orders of magnitude from sample to sample. **Zero-inflation** means that
typically more than 70% of the values are zeros, which could be due to either
physical absence (structural zeros) or insufficient sampling effort (sampling
zeros). **Compositionality** means that a change in the absolute abundance of
one taxon will lead to apparent variations in the relative abundances of other
taxa in the same sample. If neglected, such properties may cause significant
bias in the results of DAA or other statistical tests. Therefore, several
approaches have been developed to address the unique properties of microbiome
data and provide statistically reliable results.

## Common transformation methods

Let us now summarize some commonly used transformations in microbiome data
science; further details and benchmarkings available in the
references.

 * **alr**: The additive log ratio transformation is part of a broader Aitchison
   family of transformations with 'clr' and 'rclr'. Compared to them the biggest
   difference is that it selects a single feature or component as a reference
   and expresses all other features as log-ratios relative to it.
   [@Greenacre2021] provides guidance on choosing an appropriate reference
   feature.

 * **clr**: Centered log ratio transformation [@Aitchison1986] is used to
   reduce data skewness and compositionality bias in relative
   abundances, while bringing the data to the logarithmic scale. This
   transformation is frequently applied in microbial ecology as it may enhance
   comparability of relative differences between samples [@Gloor2017]. However,
   the resulting transformed values are difficult to interpret directly, and
   this transformation only applies to positive values. Usual solution for
   for making values non-zero is to add `pseudocount`, which adds another type
   of bias in the data (see 'rclr' below).

 * **hellinger**: Hellinger transformation is equal to the square root of
   relative abundances. This ecological transformation can be useful
   if we are interested in changes in relative abundances.

 * **log, log2, log10**: Logarithmic transformations, used e.g. to
   reduce data skewness. With compositional data, the `clr` (or `rclr`)
   transformation is often preferred.

 * **pa**: Presence/Absence transformation ignores abundances and only
   indicates whether the given feature is detected above the given
   threshold (default: 0). This simple transformation is relatively
   widely used in ecological research. It has shown good performance
   in microbiome-based classification performance [@Giliberti2022,
   Karwowska2024].

 * **rank**: Rank transformation replaces each value by its rank. Also
   see 'rrank' (relative rank transformation). This has use, for
   instance, in non-parametric statistics.
   
 * **rclr**: The robust `clr` ("rclr") is similar to regular clr (see above) but
   allows data with zeroes and avoids the need to add pseudocount
   [@Keshavan2010, @Martino2019].
   
 * **relabundance**: Relative transformation, also known as total sum
   scaling (TSS) and compositional transformation. This converts
   counts into percentages (at the scale [0, 1]) that sum up to 1. Much of the
   currently available taxonomic abundance data from
   high-throughput assays (16S, metagenomic sequencing) is
   compositional by nature, even if the data is provided as counts
   [@Gloor2017].

 * **standardize**: Standardize (or 'z-score') transformation scales data to zero
   mean and unit variance. This is used to bring features (or samples) to more
   comparable levels in terms of mean and scale of the values. This can enhance
   visualization and interpretation of the data

 * Other available transformations include Chi square ('chi.square'),
   Frequency transformation ('frequency'), and make margin sum of
   squares equal to one ('normalize')

Transformations on abundance assays can be performed with
`mia::transformAssay()`, keeping both the original and the transformed
assay(s). The transformed abundance assay is then stored back to the
'assays' slot in the data object. The function applies sample-wise or
column-wise transformation when MARGIN = 'cols', feature-wise or
row-wise transformation when MARGIN = 'rows'.
A complete list of available transformations and parameters, is
available in the function
[help](https://microbiome.github.io/mia/reference/transformAssay.html).

::: {.callout-important}

`Pseudocount` is a small non-negative value added to the normalized data to
avoid taking the logarithm of zero. It's value can have a significant impact
on the results when applying 
a logarithm transformation to normalized data, as the logarithm transformation
is a nonlinear operation that can fundamentally change the data distribution
[@Costea2014].

`Pseudocount` should be chosen consistently across all normalization methods
being compared, for example, by setting it to a value smaller than the minimum
abundance value before transformation. Some tools, like ancombc2, take into
account the effect of the `pseudocount` by performing sensitivity tests using
multiple pseudocount values. See [@sec-differential-abundance].

:::

## Transformations in practice

Below, we apply relative transformation to counts table.

```{r}
#| label: transf1

# Load example data
library(mia)
data("Tengeler2020")
tse <- Tengeler2020

# Transform counts assay to relative abundances
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")
```

Get the values in the resulting assay, and view some of the first
entries of it with the `head` command.

```{r}
#| label: transf2

assay(tse, "relabundance") |> head()
```

In 'pa' transformation, abundance table is converted to
presence/absence table that ignores abundances and only indicates
whether the given feature is detected. This simple transformation is
relatively widely used in ecological research. It has shown good
performance in microbiome-based classification performance
[@Giliberti2022, Karwowska2024].

```{r}
#| label: transf3

# Here, assay.type is not explicitly specified, however, it is good practice
# to do so. By default. the function uses the assay named "counts" for the
# transformation.
tse <- transformAssay(tse, method = "pa")
assay(tse, "pa") |> head()
```

You can now view the entire list of abundance assays in your data object with:

```{r}
#| label: transf4

assays(tse)
```

A common question is whether the centered log-ratio (clr) transformation should
be applied directly to raw counts or if a prior transformation, such as
conversion to relative abundances, is necessary.

In theory, the clr transformation is scale-invariant, meaning it does not matter
whether it is applied to raw or relative abundances, as long as the
relative scale of abundances remains the same. However, in practice, there are
some differences due to the introduction of a pseudocount, which can introduce
bias.

There is no single correct answer, but the following considerations may help:

- Data imputation should typically be applied to raw abundances, regardless of
the microbial profiling pipeline used or whether the obtained abundances are
counts or relative abundances.

- Once a pseudocount has been added, it makes no difference whether one first
converts to relative abundances before applying clr or applies clr directly to
the adjusted counts.

- Since applying clr directly to raw counts is the simpler approach, it is
generally recommended.

- One might also consider using robust clr instead.

```{r}
#| label: transf5

tse <- transformAssay(
    x = tse,
    assay.type = "counts",
    method = "clr",
    pseudocount = TRUE,
    name = "clr"
)
```

To incorporate phylogenetic information, one can apply the phylogenetic
isometric log-ratio (PhILR) transformation [@Silverman2017]. Unlike standard
transformations, PhILR accounts for the genetic relationships between taxa.
This is important because closely related species often share similar
properties, which traditional transformations fail to capture.

```{r}
#| label: transf6

tse <- transformAssay(tse, method = "philr", MARGIN = 1L, pseudocount = TRUE)
```

Unlike other transformations, PhILR outputs a table where rows represent nodes
of phylogeny. These new features do not match with features of `TreeSE` which
is why this new dataset is stored into `altExp`.

```{r}
#| label: transf7

altExp(tse, "philr")
```

::: {.callout-tip}
## Summary

Microbiome data is characterized by the following features:

- Compositional
- High variability
- Zero-inflated

[OSCA book](https://bioconductor.org/books/3.18/OSCA.basic/normalization.html#normalization-transformation) provides additional information on normalization from the perspective of single-cell analysis.

:::

::: {.callout-tip icon=false}
## Exercises

**Goal:** The goal is to learn how to apply different transformations.

**Exercise 1: Transform data**

1. Load any of the example datasets mentioned in [@sec-example-data].

2. Visualize counts with a histogram. Describe the data distribution. Is there
lots of zeroes?

3. Transform the counts assay into relative abundances and store it into the
`TreeSE` as an assay named `relabund`.

3. Similarly, perform a CLR transformation on the counts assay with a
`pseudocount` of 1 and add it to the `TreeSE` as a new assay.

4. List the available assays by name.

5. Visualize the CLR-transformed data with a histogram. Compare the distribution
with distribution of counts data.

6. Access the CLR-assay and store it to variable. Select a subset
of its first 100 features and 10 samples, and print the abundance table. Explore
the data.

6. Agglomerate the data with `agglomerateByRanks()` Transform data now with
`altexp = altExpNames(tse)`.

6. If the data has phylogenetic tree, perform the phILR transformation. Where
the transformed data was stored? Compare the feature names with original data.
Why the names differ?

Useful functions:

`data()`, `plotHistogram()`, `transformAssay()`, `assayNames()`, `assay()`, `agglomerateByRanks()`, `altExp()`, `rownames()`

:::
