# Transformation {#sec-assay-transform}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

Data transformations are common in (microbial) ecology [@Legendre2001].
They are used to mitigate technical biases in the data, to obtain more 
interpretable values, to enhance the comparability of samples/features or
to make data compatible with the assumptions of certain statistical methods.

Examples include transforming feature counts into relative abundances 
(i.e., "normalising as proportions"), or with compositionality-aware
transformations such as the centered log-ratio transformation (clr). 

## Characteristics of microbiome data to inform data transformations {#sec-stat-challenges}

Transformations are important in working with microbiome data due to 
various unique characteristics of sequencing data. 
For example, the above mentioned examples of feature transformations are commonly 
used with microbiome data following the inherently proportional nature of sequencing reads: 
Due to the nature of sequencing technology, read counts of taxa do not represent 
real counts of microbes in the original sample, and raw counts of taxa per sample 
are thus not comparable between samples. 

This **Compositionality** means that a change in the absolute abundance of
one taxon will lead to apparent variations in the relative abundances of other
taxa in the same sample. If neglected, such properties may cause significant
bias in the results of statistical tests. The above-mentioned transformation processes 
are one method developed to overcome these issues by making taxon abundances comparable
across samples. 

In addition to this compositionality, other special characteristics of microbiome 
sequencing data are high variability and zero-inflation.
**High variability** means that abundance of taxa often varies by several
orders of magnitude from sample to sample. **Zero-inflation** means that
typically more than 70% of the taxa-per-sample values are zeros, which could be due to either
physical absence (structural zeros) or insufficient sequencing coverage (sampling
zeros).

## Common transformation methods

Let us now summarize some commonly used transformations in microbiome data
science; further details and benchmarkings are available in the
references.

 * 'relabundance' Relative transformation, also known as normalising as proportions, 
   total sum  scaling (TSS) and compositional transformation. This converts
   counts into proportions (at the scale [0, 1]) that sum up to
   1. Much of the currently available taxonomic abundance data from
   high-throughput assays (16S, metagenomic sequencing) is
   compositional by nature, even if the data is provided as "counts"
   [@Gloor2017].

 * 'clr' Centered log ratio transformation [@Aitchison1986] is used to
   reduce data skewness and compositionality bias in these relative
   abundances, while bringing the data onto a logarithmic scale. This
   transformation is frequently applied in microbial ecology
   [@Gloor2017]. However, clr transformation can only be applied to positive
   values, not zeros. Usual solution to this is to add `pseudocounts` (See below). 
  `pseudocount` is a small number (e.g., 1), that will replace the zero for all taxa 
   that have zero count in a sample. However, this adds another type of bias in the data, 
   as true taxon absences are not taken in account. This is mitigated by the method of robust `clr` 
   transformation ('rclr'). The robust `clr` is similar to regular clr but allows 
   data with zeroes and avoids the need to add pseudocount [@Keshavan2010, @Martino2019]. 

   While the resulting values from these transformations are hard to interpret
   directly, this transformation does enhance comparability of relative
   abundance differences between samples. It is part of a broader Aitchison
   family of transformations, including also additive log ratio transformation
   (`alr'), which is is also available as a method here. 

  * 'pa' Presence/Absence transformation ignores abundances and only
   indicates whether the given feature is detected in a sample above a given
   threshold (default: 0). This simple transformation is relatively widely used in 
   ecological research, whenever we have reasons to be interested in which taxa are 
   present in which samples more so than their abundances. An example of such case
   is the process of inferring microbial ecological association networks from covariation 
   of taxa across samples. Here, microbes that are rarely seen in the same samples are 
   expected to have a negative ecological association (either competitive exclusion with 
   each other of different ecological niche preferences). Conversely, looking for negative 
   correlations in their relative abundances would not be a meaningful measure of 
   ecological associations.”

 * 'standardize' Standardize (or 'z-score') transformation scales data to zero
   mean and unit variance. This is used to bring features (or samples) to more
   comparable levels in terms of mean and scale of the values. This can enhance
   visualization and interpretation of the data

* 'log', 'log2', 'log10' Logarithmic transformations, used e.g. to
   reduce data skewness. With compositional data, the `clr` (or `rclr`)
   transformation is often preferred.

 * 'hellinger' Hellinger transformation is equal to the square root of
   relative abundances. This ecological transformation can be useful
   if we are interested in changes in relative abundances.

 * 'rank' Rank transformation replaces each feature abundance value by its rank. 
   See also 'rrank' (relative rank transformation). This has use, for
   instance, in non-parametric statistics.

 * Other available transformations include Chi square ('chi.square'),
   Frequency transformation ('frequency'), and Make margin sum of
   squares equal to one ('normalize')

Transformations on abundance assays can be performed with
`mia::transformAssay()`, keeping both the original and the transformed
assay(s) in the data object. The function applies sample-wise or
column-wise transformation when MARGIN = 'cols', feature-wise or
row-wise transformation when MARGIN = 'rows'.
A complete list of available transformations and parameters, is
available in the function
[help](https://microbiome.github.io/mia/reference/transformAssay.html).

::: {.callout-important}

`Pseudocount` is a small non-negative value added to the normalized data to
avoid taking the logarithm of zero. It's value can have a significant impact
on the results when applying a logarithm transformation to normalized data, 
as the logarithm transformation is a nonlinear operation that can fundamentally 
change the data distribution
[@Costea2014].


`Pseudocount` should be chosen consistently across all normalization methods
being compared, for example, by setting it to a value smaller than the minimum
abundance value before transformation. Some tools, like ancombc2, take into
account the effect of the `pseudocount` by performing sensitivity tests using
multiple pseudocount values. See [@sec-differential-abundance].

:::

## Transformations in practice

```{r}
# Load example data
library(mia)
data("GlobalPatterns", package = "mia")
tse <- GlobalPatterns

# Transform "counts" assay to relative abundances ("relabundance")
tse <- transformAssay(tse, assay.type = "counts", method = "relabundance")

# Transform "counts" to "clr", using pseudocount if necessary; name
# the resulting assay to "clr".
tse <- transformAssay(
    x = tse, assay.type = "counts", method = "clr", pseudocount = TRUE,
    name = "clr")

```

Get the values in the resulting assay, and view some of the first
entries of it with the `head` command.

```{r}
assay(tse, "clr") |> head()
```

In 'pa' transformation, abundance table is converted to
presence/absence table that ignores abundances and only indicates
whether the given feature is detected in the sample. 

```{r}
# Here, `assay.type` is not explicitly specified and the function uses the "counts" assay for the transformation.
tse <- transformAssay(tse, method = "pa")
assay(tse, "pa") |> head()
```

You can now view the entire list of abundance assays in your data object with:

```{r}
assays(tse)
```

::: {.callout-tip}
## Summary

Microbiome data is characterized by the following features:

- Compositionality
- High variability
- Zero-inflation

[OSCA book](https://bioconductor.org/books/3.18/OSCA.basic/normalization.html#normalization-transformation) provides additional information on normalization from the perspective of single-cell analysis.

:::

::: {.callout-tip icon=false}
## Exercises

**Goal:** The goal is to learn how to apply different transformations.

**Exercise 1: Transform data**

1. Load any of the example datasets mentioned in [@sec-example-data].

2. Visualize counts with a histogram. Describe the data distribution. Is there
lots of zeroes?

3. Transform the counts assay into relative abundances and store it into the
`TreeSE` as an assay named `relabund`.

3. Similarly, perform a CLR transformation on the counts assay with a
`pseudocount` of 1 and add it to the `TreeSE` as a new assay.

4. List the available assays by name.

5. Visualize the CLR-transformed data with a histogram. Compare the distribution
with distribution of counts data.

6. Access the CLR-assay and store it to variable. Select a subset
of its first 100 features and 10 samples, and print the abundance table. Explore
the data.

6. Agglomerate the data with `agglomerateByRanks()` Transform data now with
`altexp = altExpNames(tse)`.

6. If the data has phylogenetic tree, perform the phILR transformation. Where
the transformed data was stored? Compare the feature names with original data.
Why the names differ?

Useful functions:

`data()`, `plotHistogram()`, `transformAssay()`, `assayNames()`, `assay()`, `agglomerateByRanks()`, `altExp()`, `rownames()`

:::
