# Data wrangling {#sec-datawrangling}

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

This chapter introduces several essential techniques for preparing data for
analysis. These techniques include splitting data, modifying data, 
and converting data to a `data.frame.` Additionally, it explains how to merge 
multiple `SummarizedExperiment` objects when needed. For a basic understanding
of  `TreeSE` please refer to [@sec-containers]

## Splitting {#sec-splitting}

You can split the data based on variables by using the functions 
`agglomerateByRanks()` and `splitOn()`. The former is detailed in
[@sec-agglomeration].

If you want to split the data based on a variable other than taxonomic rank, use 
`splitOn()`. It works for row-wise and column-wise splitting. Splitting the data
may be useful, for example, if you want to analyze data from different cohorts
separately.

The following example demonstrates how to identify the most abundant taxonomic
features for each sample type. Given that some samples originate from different
origin, we expect differences between sample types. While `getTop()` can be used
to determine the most abundant features across the entire dataset, this
approach would obscure features that are specific to individual sample types.

```{r}
#| label: splitOn

library(mia)
library(knitr)

data("GlobalPatterns")
tse <- GlobalPatterns

# Agglomerate to phyla
tse <- agglomerateByRank(tse, rank = "Phylum")

# Split data based on sample type
tse_list <- splitOn(tse, by = "samples", group = "SampleType")

# Loop over the list of TreeSEs, and get top features for each
top_taxa <- sapply(tse_list, getTop)

top_taxa |> kable()
```

## Add or modify variables {#sec-add-or-modify-data}

The information contained by the `colData` of a `TreeSE` can be added and/or 
modified by accessing the desired variables. You might want to add or modify 
this data to include new variables or update existing ones, which can be essential 
for ensuring that all relevant metadata is available for subsequent analyses.

```{r}
#| label: modify_coldata

data("GlobalPatterns")
tse <- GlobalPatterns

# modify the Description entries
colData(tse)$Description <- paste(
    colData(tse)$Description, "modified description")

# view modified variable
tse$Description |> head()
```

New information can be added to the experiment by creating a new variable.

```{r}
#| label: add-coldata

# simulate new data
new_data <- runif(ncol(tse))

# store new data as new variable in colData
colData(tse)$NewVariable <- new_data

# view new variable
tse$NewVariable |> head()
```

Alternatively, you can add whole table by merging it with existing `colData`.

```{r}
#| label: add-coldata2

# simulate new data
new_data <- data.frame(var1 = runif(ncol(tse)), var2 = runif(ncol(tse)))
rownames(new_data) <- colnames(tse)

# Combine existing data with new data
colData(tse) <- cbind(colData(tse), new_data)
```

Similar steps can also be applied to `rowData`. If you have an assay whose
rows and columns aling with the existing ones, you can add the assay easily
to the `TreeSE` object.

Here we add an assay that has random numbers but in real life these steps
might come handy after you have transformed the data with custom transformation
that cannot be found from `mia`.

```{r}
#| label: add_assay

# Create a matrix with random values
mat <- rnorm(ncol(tse)*nrow(tse), 0, 1)
mat <- matrix(mat, ncol = ncol(tse), nrow = nrow(tse))
# Add matrix to tse
assay(tse, "random", withDimnames = FALSE) <- mat

assayNames(tse)
```

Now we can see that the `TreeSE` object has now an additional assay called
"random". When adding new samples or
features to your existing dataset, you can use `cbind()` to combine columns for
new features or `rbind()` to add rows for new samples.

```{r}
#| label: combine-cols

tse2 <- cbind(tse, tse)
tse2
```

However, the aforementioned functions assume that the rows align correctly when
combining columns, and vice versa. In practice, this is often not the case;
for example, samples may have different feature sets. In such situations, using
a merging approach is the appropriate method.

## Merge data

`mia` package has `mergeSEs()` function that merges multiple
`SummarizedExperiment` objects. For example, it is possible to combine
multiple `TreeSE` objects which each includes one sample.

`mergeSEs()` works much like standard joining operations. It combines rows and 
columns and allows you to specify the merging method.

```{r}
#| label: merge1

# Take subsets for demonstration purposes
tse1 <- tse[, 1]
tse2 <- tse[, 2]
tse3 <- tse[, 3]
tse4 <- tse[1:100, 4]
```

```{r}
#| label: merge2

# With inner join, we want to include all shared rows. When using mergeSEs
# function all samples are always preserved.
tse <- mergeSEs(list(tse1, tse2, tse3, tse4), join = "inner")
tse
```

```{r}
#| label: merge3

# Left join preserves all rows of the 1st object
tse <- mergeSEs(tse1, tse4, missing.values = 0, join = "left")
tse
```

## Melting data

For several custom analysis and visualization packages, such as those from
`tidyverse`, the `SE` data can be converted to a long `data.frame` format with 
`meltSE()`.    

```{r}
#| label: melt-data

library(knitr)

# Melt SE object
molten_tse <- meltSE(
    tse,
    assay.type = "counts",
    add.row = TRUE,
    add.col = TRUE,
    )

molten_tse |> head() |> kable()
```

For `MultiAssayExperiment` data, you can utilize instead `longForm()`.

```{r}
#| label: melt-mae-data

data(HintikkaXOData)
mae <- HintikkaXOData

molten_mae <- longForm(
    mae,
    i = c("counts", "nmr", "signals"),
    colDataCols = c("Rat", "Diet")
    )

molten_mae |> head() |> kable()
```

## Tidy R programming

The tidy paradigm, first introduced in the tidyverse, promotes an intuitive and
easy-to-learn coding style, which has made it popular among users. _tidyomics_
bridges the gap between Bioconductor’s `SummarizedExperiment` ecosystem and the
_tidyverse_ [@Hutchison2024].

The package `TidySingleCellExperiment` can be used to manage data in `TreeSE`.
For instance, we can easily view the object as `tibble` abstraction, familiar
from _tidyverse_.

```{r}
#| label: tidy1

data("GlobalPatterns")
tse <- GlobalPatterns

library(tidySingleCellExperiment)

tse
```

By utilizing tidy programming, we could then effortlessly manipulate the data.
For instance, we can calculate the mean library size in each
sample type. This is done by first calculating library size for each sample and
then summarizing total counts in each sample type; all done by using tidy
paradigm. For more information on library size, see [@sec-quality-control].

```{r}
#| label: tidy2

tse %>%
    join_features(features = rownames(tse)) %>%
    group_by(.cell) %>%
    mutate(total_counts = sum(.abundance_counts)) %>%
    group_by(SampleType) %>%
    summarise(total_counts = mean(total_counts))
```

We can then filter samples and leverage `ggplot2` for plotting. In the
example below, we select soil samples and visualize their abundances with a
boxplot. Of course, these tidy commands can be combined with `mia` tools. Before
plotting, we apply CLR transformation by using `trasformAssay()`
(see more on transformation from [@sec-assay-transform]).

```{r}
#| label: tidy3

library(ggplot2)

tse %>%
    filter(SampleType == "Soil") %>%
    transformAssay(method = "clr", pseudocount = TRUE) %>%
    join_features(features = rownames(tse)) %>%
    ggplot(aes(x = .cell, y = .abundance_clr, fill = .cell)) +
    geom_boxplot()
```

As demonstrated, tidy R programming can be effectively used to manage `TreeSE`
objects. Please refer to
[tidySummarizedExperiment](https://www.bioconductor.org/packages/release/bioc/vignettes/tidySummarizedExperiment/inst/doc/introduction.html)
and
[tidySingleCellExperiment](https://www.bioconductor.org/packages/release/bioc/vignettes/tidySingleCellExperiment/inst/doc/introduction.html)
vignettes for more examples.

::: {.callout-tip icon=false}
## Exercises

**Goal:** The goal is to learn how to work with multiple datasets by merging
them and splitting dataset into list of `TreeSE`s.

**Exercise 1: Working with multiple datasets**

1. Load any of the example datasets mentioned in [@sec-example-data].

2. Explore sample metadata.

3. Add arbitrary groups to `colData`. For instance, you create groups like this:
`sample(c("group1", "group2"), size = ncol(tse), replace = TRUE)`.

4. Explore the group distribution with a bar plot.

5. Split the data based on groups. The result should be a list containing
each group in separate `TreeSE`.

6. For each `TreeSE`, explore the grouping variable in `colData`. Does each
`TreeSE` include only one unique value?

7. Calculate the total number of counts in each group.
    - Loop through elements of the list.
    - For single `TreeSE`, retrieve `assay`, sum all the values, and add
    the calculated sum to `colData` (all the samples in certain group should
    have same value).
    - The result should be a list of `TreeSE`s with additional sample metadata
    column.

8. Combine the list back to single `TreeSE`.

9. Load another dataset from [@sec-example-data] and merge it with the one
You've been using. Explore the data.

Useful functions:

`data()`, `colData()`, `sample()`, `plotBarplot()`, `splitOn()`, `lapply()`,
`assay()`, `sum()`, `unsplitOn()` `mergeSEs()`

:::
